|

### 2021-12-02 01-Chp-Background.Rmd
End of day:

```{block type='savequote', include=knitr::is_latex_output(), quote_author='(ref:dara-quote)', echo = TRUE}
Considering everything, it seems we are dealing
here with a special illness.
```
(ref:dara-quote) --- Alois Alzheimer, 1907 [@alzheimer1907; @stelzmann1995]

@@

<!-- TODO Contributor statements -->

## Lay summary

Around 850,000 people in the UK live with dementia, and by 2040, nearly twice as many will have the condition. Despite many promising candidates, no cure for dementia currently exists, meaning the focus is on finding ways to prevent the condition. The best way to do this is to find risk factors (characteristics that influence a person's chance of developing a disease) for dementia that we can easily change. Avoiding a risk factor does not guarantee that a person will not develop dementia but makes it less likely. A key risk factor for dementia may be the levels of lipids (fatty substances such as cholesterol) in a person's blood, though not all existing research agrees. The aim of this thesis is to use all available evidence to assess whether blood lipids levels are in fact a risk factor for dementia.

@@

@@

@@

Defined by the Diagnostic and Statistical Manual of Mental Disorders as a "major neurocognitive disorder", dementia is a progressive disease which impairs cognitive functions including speech, memory and executive reasoning.[@edition2013] At advanced stage, the condition causes severe behavioral and personality changes,[@cerejeira2012] cumulating in reduced motor control that affects patients ability to swallow or breathe.[@kumar2013] The condition has several distinct underlying causes, including Alzheimer's disease and vascular dementia.[@burns2009]

Alzheimer's disease is the most common cause of dementia, accounting for approximately 60-80% of cases. Characterised by substantial cognitive impairment and difficulty with high level executive function to the extent that it interferes with, it is an insidious disease, within initial onset thought to occur up to 15 years prior to symptomatic presentation.[@robinson2015] Much remains unknown about Alzheimer's pathogenesis, despite research implicating the "amyloid hypothesis",[@robinson2015] as a potential mechanism of disease. Under this hypothesis, the build-up of amlyoid plaques (composed mainly of amlyoid-$\beta$ peptide) and neurofibrillary tangles (composed mainly of tau protein) triggers a range of physiological changes, including inflammation and cell death, that result in cognitive impairment.[@robinson2015]

@@

@@

One of the most commonly used criteria are those found in the Diagnostic and Statistical Manual of Mental Disorders (DSM) criteria (Table \@ref(tab:diagnosticCriteria-table)).[@edition2013] These criteria are outlined in Table \@ref(tab:diagnosticCriteria-table), and form the broad definition of a dementia diagnoses, supported by a detailed patient history, evidence from carers and family members, and objective assessments of cognitive ability using neurocognitive tests.

@@

<!-- ### Differing risk by sex -->

<!-- Transparency guidelines that suggest sex should eb treated as biological variable and so results should be presented stratified t -->

<!-- Rate of dementia/AD differs in men and women <!--- Though is this just due to the different lifestyles and life expectancies --->

<!-- Previous studies have hinted that the relationship is dependent on individual-level characteristics such as age and sex [16,28]. -->

@@

Dementia is quickly becoming a critically important public health issue. Despite the age-specific incidence and prevalence of dementia remaining relatively constant over time,[@prince2016] an ageing population is set to create a dementia epidemic, particularly in Westernised countries.[@flier2005] While approximately 525,000 patients have received a dementia diagnosis, the true number of people currently living with dementia in the UK is thought to be closer to 850,000, with this figure expected to double by 2040.[@baker2019] Globally, the prevalence of dementia is expected to reach 75 million by 2030.[@prince2016] Dementia is the leading cause of death in the UK, and the only one without a proven cure.

Dementia also has a substantial economic impact. In 2015, the estimated total cost of dementia in England was £24.2 billion. Health care costs alone were £3.8 billion,[@wittenberg2019] Globally, the cost of dementia care is expected to rise to $1tr by 2030.[@prince2014]

As such, the urgent need to reduce the burden of dementia, both at the personal and system level, is clear.

@@

Developing treatments for dementia is regularly deemed to be one of the hardest markets in the pharmaceutical world, with trials of seemingly promising therapeutics being regularly abandoned due to futility.[@cummings2020]. At present, there are no known curative treatments for dementia, regardless of the underlying cause, though several available therapeutics can help alleviate the symptoms of Alzheimer's disease.

The most common of these are acetylcholinesterase (ACE) inhibitors, which inhibit the degradation of the neurotransmitter acetylcholine by competitively binding the ACE enzyme. Acetylcholine plays a key role in controlling the cholingeric synapses, which are highly concentrated in regions of the brain (such as the neocortex) that control higher level brain functions such as memory and attention.[@hampel2018] Commonly prescribed ACE inhibitors include donepezil and galantamine.[@pariente2008] ACE inhibitors increase the availability of the neurotransmitter, and has shown clinical effect is easing the behavioural and memory-related symptoms of Alzheimer's disease.[@marucci2020] ACE inhibitors represent only a stop-gap treatment, treating the symptoms rather than the underlying pathology which may continue to progress.[@francis2010]

@@

In addition to the individual fractions, total serum cholesterol (TC) is a commonly-used summary measure to estimate the total amount of lipid present in the blood. The measure is derived from measurements of the individual HDL-c, LDL-c and TG levels using the Friedwald formula:[@friedewald1972]

@@

where $k$ is 0.20 if measurements are in milligrams per decilitre (_mg/dl_) and 0.45 if measured in millimole per litre (_mmol/l_).

Widely-used ranges for the acceptable levels of different types of lipids are based on the National Cholesterol Education Program (NCEP)[@national2002third], and are outlined in Table \@ref(tab:lipidLevels-table).

@@

```{r lipidLevels-table, message=FALSE, results="asis", echo = FALSE}
```
<!----------------------------------------------------------------------------->

Elevated LDL-c in the bloodstream, a condition also known as hypercholesterolaemia or hyperlipidaemia,[@nelson2013] can lead to atherosclerosis,[@libby2019] the build-up of fatty deposits in the blood vessels. These deposits constrict blood flow and can lead to vascular complications. Alternatively, part of the deposit can detach from the artery walls, forming a clot that can lead to a heart attach or stroke.[@libby2019] Globally, the prevalence of elevated cholesterol was estimated by the World Health Organization to be approximately 40%.

@@

<!----------------------------------------------------------------------------->
<!-- TODO Review the look and feel of this table -->

@@

There are several other interventions that can be used to modify a persons lipid profile, which each acting in slightly different ways (Table \@ref(tab)). However, in general, these treatments are either used as adjunct (additional) treatments with statins therapy or are used in situations where statins are contra-indicated or not tolerated.

@@

<!----------------------------------------------------------------------------->
### Basic science {#intro-basic-science}

A role for lipids in the aetiology of the dementia is supported by both genetic linkage studies and functional cell biology studies. The generation of the amyloid plaques found in the brains of Alzheimer's patients is cholesterol dependent,,[@burns2003; @mizuno1999] while the most established genetic risk factor for late-onset dementia, apolipoprotein E (ApoE), is involved in cerebral cholesterol transport. Several other genes involved in cholesterol transport have also been found to be associated with increased AD susceptibility.[@beecham2014; @harold2009; @meng2007]

@@

<!----------------------------------------------------------------------------->
### Observational studies

By far the largest source of evidence on the relationship between comes from observational designs. Several studies have examined the relationships between concentrations of serum lipids (total cholesterol (TC), low density lipoprotein cholesterol (LDL-c), high density lipoprotein cholesterol (HDL-c) and triglycerides) and both Alzheimer's disease and vascular dementia and reported extremely varied results. In some studies, a high serum cholesterol concentration has been found to be associated with an increase in susceptibility to AD,[@kivipelto2002; @kivipelto2005; @schilling2017; @solomon2009; @whitmer2005] however others have shown no association,[@li2005; @mainous2005; @mielke2010; @tan2003] or a reduced susceptibility [@mielke2005; @reitz2004]. With regards vascular dementia, decreased levels of HDL-c appear to be associated with increased risk [@reitz2004], while for LDL-c, studies have reported both positive and negative associations.[@reitz2004; @moroney1999]

<!-- Several previous systematic review of observational studies examining the effect of lipids[@anstey2015] and lipid-regulating agents[@chu2018;@poly2020] on dementia outcomes have been performed. However, these reviews have several limitations. Many did not consider grey literature sources (see Section \@ref(diverse-sources-preprints)). Additionally, many of the reviews of observational studies did not perform any risk-of-bias assessment[@chu2018] or used an outdated assessment tool.[@anstey2015;@poly2020c] -->

@@

In terms of the central research of this thesis, RCTs of statin therapy can be used to provide indirect evidence for the effect of reducing blood LDL-c levels on dementia risk.

However, RCTs may be infeasible if the outcome of interest is one with a long prodomal period, such as dementia (see Section \@ref(underlying-pathologies)), as they would require extremely long and costly follow-up.[@ritchie2015] It is no surprise then that the two previous trials providing evidence on the effect of statins on dementia risk, identified by a recent Cochrane review,[@mcguinness2016] are in fact trials of statins for the prevention of coronary related outcomes.

While being widely cited, these studies have major limitations that reduce their utility as a source of evidence on the effect of statin treatment on in assessing the impact of lipid-lowering treatment on dementia risk. Firstly, there was no clinical cognitive evaluation of patients to determine a dementia outcome. One of the trials, the Prospective Study of Pravastatin in the Elderly (PROSPER) trial,[@trompet2010] reported not on dementia outcomes but on the change in cognitive scores over a mean of 3.2 years. As highlighted in Section \@ref(diagnostic-criteria), a "change in score" alone is insufficient to diagnose a dementia outcome. The second trial, the Medical Research Council/British Health Foundation Protection Study,[@heartprotectionstudycollaborativegroup2002] found no effect of simvastatin on dementia (`r estimate(1, 0.61, 1.65)`), but did not report how the outcome was assessed/recorded within the trial.

Additionally, the two trials did not make any effort to assign an underlying pathology to each case, instead reporting an all-cause dementia outcome. As discussed in Section \@ref(underlying-pathologies), the different underlying pathology of dementia have different mechanisms of action, and so it is not gauranteed that the effect of statins would be consistent across them.

Both trials were also limited by the relatively short follow-up period examined, expected when the primary outcome of the trials were coronary related conditions rather than dementia.[@heartprotectionstudycollaborativegroup2002; @trompet2010] The PROSPER trial had a mean follow-up of 3.2 years, while the MRC/BHF Protection Study estimated risk at 5 years of follow-up. Given the long lag time between non-symptomatic onset of dementia and clinical presentation, it is likely that these durations are insufficient to fully capture the onset of dementia. Finally, as they included only patients at high vascular risk, their generalisability to other settings is limited.[@mcguinness2016]

@@

Newer methodological approaches, such as Mendelian randomisation (MR),[@daveysmith2014] have also been used to examine the effect of varying lipid levels on dementia risk in an effort to combat the risk of reverse causation and residual confounding inherent to observational studies.[@greenland2000] In brief, MR uses genetic variants that are both strongly associated with the exposure of interest and are independent from potential confounders to strengthen causal inference.[@daveysmith2014] The analytic method relies on several assumptions about the instrumental variable (IV),[@davies2018] namely that:

@@

Recent MR studies indicated that genetically determined low levels of LDL-c may cause a reduction in AD risk.[@larsson2017; @ostergaard2015] However, the effect was attenuated in sensitivity analysis that exclude the region surrounding the ApoE gene, the strongest known risk factor for Alzheimer's disease.[@kim2009] Inclusion of ApoE4 variants invalidates the exclusion restriction criteria (Assumption 3, above), as the risk reduction observed may be driven by variants in this region via a pathway independent of lipid levels. This was supported by further MR studies where _ApoE4_ variants were intentionally excluded.[@benn2017]

Despite the increasing number of MR studies examining this topic, no systematic review of this study design as a source of evidence has been performed.

@@

<!-- TODO Consider inverting the order so that triangulation comes last - as it does in the thesis as a whole! -->

Evidence synthesis is the process of finding and integrating information from several sources to examine a research question.[@donnelly2018] A common tyoe of evidence synthesis is a systematic review, either with or without a meta-analysis.[@chandler2019chapter]

@@

* Inclusion of preprints
* Triangulation across evidence sources
* Individual patient data meta-analysis

@@

<!-- TODO Incorporate Julian's comments below -->

<!-- REVIEW [JH] I think, on general reflection, that portraying this as an issue around “grey literature” may not be particularly useful. Preprints are arguably some of the least “grey” of the very ambiguously defined family of grey sources, such that some definitions of “grey” might not include them.
It seems to me that preprints are such a big and important part of the landscape now that you can just portray the methodology as dealing with preprints rather than including grey literature.
Interesting debate as to whether preprints are grey literature, given that many of them are increasingly becoming findable in electronic bibliographic databases. We have no clear and widely used definition of grey literature.  -->

The importance of including grey (or gray) literature in systematic reviews is widely acknowledged. Meta-research studies have demonstrated that systematic reviews excluding grey literature sources overestimate the effect of interventions.[@conn2003; @mcauley2000; @hopewell2007] Common, well-accepted forms of grey literature include conference abstracts and theses.[@lefebvre2019searching]

A important developing source of grey literature are preprints. Defined by the Committee on Publication Ethics (COPE) as 'scholarly manuscript[s] posted by the author(s) in an openly accessible platform, usually before or in parallel with the peer review process'[@committeeonpublicationethicscope2018], preprints serve several purposes. They are used to establish primacy when submitting to a journal where the peer-review process may take several months,[@vale2016] to rapidly disseminate research findings, as occurred during the COVID-19 pandemic,[@fraser2020preprinting] and to make available publications that may not have been accepted elsewhere in an attempt to combat publication bias or the "file-drawer" effect.[@rosenthal1979]

@@

In this thesis, preprints are considered an important source of evidence, in contrast to previous reviews on this topic. However, as with many sources of grey literature,[@mahood2014] there are several logistical issues with carrying out systematic searches in preprint repositories. As such, to enable the inclusion of preprints in the systematic review described in Chapters \@ref(sys-rev-methods-heading), a new tool addressing these issues is presented in Chapter \@ref(sys-rev-tools).

@@

As illustrated in Section \@ref(evidence-association), several diverse epidemiological methods have been used to examine the effect of varying blood lipid levels on dementia risk. However, each method is limited by its own biases. Aetiological triangulation is a developing evidence synthesis method that seeks to exploit these inherent differences in study design, and as a result, in biases.[@lawlor2016] If several sources of evidence are available and point towards identical conclusions about an exposure-outcome relationship, and these sources are at risk of unrelated biases, this strengthens our confidence in the result. The ideal scenario is where predicted sources of bias are likely to be in competing directions, strengthen the effect of the exposure and the other to attenuate it.[@lawlor2016] As such, triangulating these results can provides us a middle-ground between the competing directions of bias. A triangulation approach can also prove useful in a prospective manner, helping to design new studies that are at risk of different sources of bias to that already available from the published literature.[@munafo2018]

This thesis seeks to apply a triangulation approach to provide the best available evidence on the effect of lipids, and lipid regulating agents, on dementia outcomes.

All existing evidence, regardless of study design, is first identified by the by the systematic review presented inChapters \@ref(sys-rev-methods-heading)/\@ref(sys-rev-results-heading). Risk-of-bias assessment using a domain-based tool is already a recommended part of the systematic review process, but is particularly important to a triangulation exercise. [@page2021; @sterne2019; @mcguinness2018] As such, a core component of the review is a comprehensive domain-based risk-of-bias assessment for all included studies.

Finally all evidence, both pre-existing and produced as part of this thesis (Chapter \@ref(cprd-analysis-heading) and \@ref(ipd-analysis-heading)), are triangulated in Chapter \@ref(dicussion-heading).

Individual patient data meta-analyses are commonly held to represent the gold standard in evidence synthesis methodology.[@riley2010; @stewart1993] IPD methods seek to obtain the raw data from each study identified in a systematic review, rather than basing the meta-analysis on summary results extracted from the literature.[@riley2010]

In the context of this thesis, if lipids are found to have a causal role in development of dementia, evidence-based preventative strategies would be best informed by identifying the types of individuals who are most likely to receive benefit from treatment with lipid-modifying agents.[@arain2009; @clayton2018; @mccartney2016] However, if primary studies do not present results stratified by covariates of interest, meta-analyses of summary-level data on this topic often have limited ability to examine research questions related to exposure-covariate interactions.[@riley2010] In terms of this thesis, patient sex is considered to be of particular interest.[@arain2009; @letenneur1999]

An IPD meta-analysis of lipid levels on dementia outcomes would overcome this limitation of summary-level data, as access to the raw data allows for an analysis that investigates these interactions.[@riley2020] This approach has the added benefit of allowing a common set of inclusion criteria and statistical model to be applied across all datasets, potentially eliminating some important sources of heterogeneity.[@stewart2002]

Despite their advantages, IPD meta-analysis are rarely performed.[@tugwell2010] Factors limiting their uptake include the increased time and effort they require when compared to a summary-level analysis, and the low success rate associated with obtaining the raw data.[@nevitt2017; @ventresca2020] The data underlying primary studies are frequently not publicly available,[@alsheikh-ali2011; @federer2018] and the availability of data "available on request from authors" declines rapidly over time.[@vines2014] Several systematic barriers to open data sharing have been identified[@vanpanhuis2014a]. Of particular concern for biomedical IPD analyses are legal issues surrounding the sharing of medical data, motivated by concerns around patient privacy.[@wartenberg2010]

I will will attempt to obtain the raw data from relevant primary studies identified by the systematic review in Chapters \@ref(sys-rev-methods-heading)/\@ref(sys-rev-results-heading). Any data obtained will be combined with that available from the DPUK portal as part of an individual participant data meta-analysis in Chapter \@ref(ipd-analysis-heading), enabling the assessment of the effect of lipids on dementia stratified by key variables such as sex.

@@

* To create a tool that allows for the inclusion of health related preprints in evidence syntheses in a systematic and reproducible manner
* To review all available evidence across multiple diverse study designs to assess the effect of lipids and lipid regulating agents on dementia risk
* To examine whether there is evidence for an effect of lipid-regulating agents on dementia and related outcomes in a large scale population-based cohort, the Clinical Practice Research Datalink (CPRD)
* To meta-analyse raw dementia-related datasets as part of a individual participant data (IPD meta-analysis) to produce evidence on exposure-covariate interactions
* To propose a generalised framework for the quantitative triangulation across diverse evidence sources

@@

<!-- TODO Ideally the two preprints for the CPRD and systematic review papers will go here -->

>>>>

```{block type='savequote', include=knitr::is_latex_output(), quote_author='(ref:alzheimer-quote)', echo = TRUE}
Alles in allem genommen haben wir hier offenbar einen eigenartigen Krankheitsprozeß vor uns.

~

Considering everything, it seems we are dealing here with a special illness.
```
(ref:alzheimer-quote) --- Alois Alzheimer, 1907 [@alzheimer1907; @stelzmann1995]

@@

::: {.laybox data-latex=""}

## Lay summary {-}

Around 850,000 people in the UK live with dementia, and by 2040, nearly twice as many will have the condition. Despite many promising candidates, no cure for dementia currently exists, meaning the focus is on finding ways to prevent the condition. The best way to do this is to find risk factors (characteristics that influence a person's chance of developing a disease) for dementia that we can easily change. Avoiding a risk factor does not guarantee that a person will not develop dementia but makes it less likely. A key risk factor for dementia may be the levels of lipids (fatty substances such as cholesterol) in a person's blood, though not all existing research agrees. The aim of this thesis is to use all available evidence to assess whether blood lipid levels are in fact a risk factor for dementia.

@@

:::

&nbsp;<!----------------------------------------------------------------------->

@@

<!--- No nb here ---------------------------------------------------------------->

@@

Defined by the Diagnostic and Statistical Manual of Mental Disorders as a "major neurocognitive disorder", dementia is a progressive disease which impairs cognitive functions including speech, memory and executive reasoning.[@edition2013] At advanced stage, the condition causes severe behavioral and personality changes,[@cerejeira2012] culminating in reduced motor control that affects patients' ability to swallow or breathe.[@kumar2013] The condition has several distinct underlying causes, including Alzheimer's disease and vascular dementia.[@burns2009]

Alzheimer's disease is the most common cause of dementia, accounting for approximately 60-80% of cases. Characterised by substantial cognitive impairment and difficulty with high level executive function, it is an insidious disease, with initial onset thought to occur up to 15 years prior to symptomatic presentation.[@robinson2015] Much remains unknown about Alzheimer's pathogenesis, despite research implicating the "amyloid hypothesis",[@robinson2015] as a potential mechanism of disease. Under this hypothesis, the build-up of amlyoid plaques (composed mainly of amlyoid-$\beta$ peptide) and neurofibrillary tangles (composed mainly of tau protein) triggers a range of physiological changes, including inflammation and cell death, that result in cognitive impairment.[@robinson2015]

@@

One of the most commonly used sets of criteria for diagnosing dementia cases are the Diagnostic and Statistical Manual of Mental Disorders (DSM) criteria (Table \@ref(tab:diagnosticCriteria-table)).[@edition2013] Outlined in Table \@ref(tab:diagnosticCriteria-table), these form the broad definition of a dementia diagnoses, and are supported by a detailed patient history, evidence from carers and family members, and objective assessments of cognitive ability using neurocognitive tests.

@@

@@

&nbsp;<!----------------------------------------------------------------------->

@@

Dementia is quickly becoming a critically important public health issue. Despite the age-specific incidence and prevalence of dementia remaining relatively constant over time,[@prince2016] an ageing population is set to create a dementia epidemic, particularly in Westernised countries.[@flier2005] While approximately 525,000 patients have received a dementia diagnosis, the true number of people currently living with dementia in the UK is thought to be closer to 850,000, with this figure expected to double by 2040.[@baker2019] Globally, the prevalence of dementia is expected to reach 75 million by 2030.[@prince2016] Dementia is a leading cause of death in the UK,[@zotero-15757] and one of the few without a proven treatment.

Dementia also has a substantial economic impact. In 2015, the estimated total cost of dementia in England was £24.2 billion. Health care costs alone were £3.8 billion, with the remainder being divided between unpaid care and social care costs.[@wittenberg2019] Globally, the cost of dementia care is expected to rise to $1tr by 2030.[@prince2014] As such, the urgent need to reduce the burden of dementia, both at the personal and system level, is clear.

&nbsp;<!----------------------------------------------------------------------->

@@

Developing treatments for dementia is regularly deemed to be one of the most challenging markets in the pharmaceutical world, with trials of seemingly promising therapeutics being regularly abandoned due to futility.[@cummings2020]. At present, there are no known curative treatments for dementia, regardless of the underlying cause, though several available therapeutics can help alleviate the symptoms of Alzheimer's disease. 
The most common of these are acetylcholinesterase (ACE) inhibitors, which inhibit the degradation of the neurotransmitter acetylcholine by competitively binding the ACE enzyme. Acetylcholine plays a key role in controlling the cholingeric synapses, which are highly concentrated in regions of the brain (such as the neocortex) that control higher level brain functions such as memory and attention.[@hampel2018] Commonly prescribed ACE inhibitors include donepezil and galantamine.[@pariente2008] ACE inhibitors increase the availability of the neurotransmitter, and have shown clinical effect is easing the behavioural and memory-related symptoms of Alzheimer's disease.[@marucci2020] ACE inhibitors represent only a stop-gap treatment, treating the symptoms rather than the underlying pathology which may continue to progress.[@francis2010]

&nbsp;<!----------------------------------------------------------------------->

@@

In addition to the individual fractions, total serum cholesterol (TC) is a commonly-used summary measure to estimate the total amount of lipid present in the blood. The level of TC is derived from measurements of the individual HDL-c, LDL-c and TG levels using the Friedwald formula:[@friedewald1972]

@@

where $k$ is 0.20 if measurements are in milligrams per decilitre (_mg/dl_) and 0.45 if measured in millimole per litre (_mmol/l_). Widely-used ranges for the acceptable levels of different types of lipids are based on the National Cholesterol Education Program (NCEP)[@national2002third], and are outlined in Table \@ref(tab:lipidLevels-table).

Elevated LDL-c in the bloodstream, a condition also known as hypercholesterolaemia or hyperlipidaemia,[@nelson2013] can lead to atherosclerosis,[@libby2019] the build-up of fatty deposits in the blood vessels. These deposits constrict blood flow and can lead to vascular complications. Alternatively, part of the deposit can detach from the artery walls, forming a clot that can lead to a heart attack or stroke.[@libby2019]

&nbsp;<!----------------------------------------------------------------------->

@@

```{r lipidLevels-table, message=FALSE, results="asis", echo = FALSE}
```
&nbsp;<!----------------------------------------------------------------------->

@@

@@

There are several other interventions that can be used to modify a persons lipid profile, each of which act via different mechanisms (Table \@ref(tab)). However, in general, these treatments are either used as adjunct (additional) treatments with statins therapy or are used in situations where statins are contra-indicated or not tolerated.

@@

&nbsp;<!----------------------------------------------------------------------->

### Basic science {#intro-basic-science}

A role for lipids in the aetiology of dementia is supported by both genetic linkage studies and functional cell biology studies. The generation of the amyloid plaques found in the brains of Alzheimer's patients is cholesterol dependent,,[@burns2003; @mizuno1999] while the most established genetic risk factor for late-onset dementia, apolipoprotein E (ApoE), is involved in cerebral cholesterol transport. Several other genes involved in cholesterol transport have also been found to be associated with increased AD susceptibility.[@beecham2014; @harold2009; @meng2007]

@@

&nbsp;<!----------------------------------------------------------------------->

### Observational studies

By far the largest source of evidence on the relationship between blood lipids and dementia outcomes comes from observational designs. Several studies have examined the relationships between concentrations of serum lipids (total cholesterol (TC), low density lipoprotein cholesterol (LDL-c), high density lipoprotein cholesterol (HDL-c) and triglycerides) and both Alzheimer's disease and vascular dementia and reported extremely varied results. In some studies, a high serum cholesterol concentration has been found to be associated with an increase in susceptibility to AD,[@kivipelto2002; @kivipelto2005; @schilling2017; @solomon2009; @whitmer2005] however others have shown no association,[@li2005; @mainous2005; @mielke2010; @tan2003] or a reduced susceptibility [@mielke2005; @reitz2004]. With regard to vascular dementia, decreased levels of HDL-c appear to be associated with increased risk [@reitz2004], while for LDL-c, studies have reported both positive and negative associations.[@reitz2004; @moroney1999]

<!-- IDEA May need summary of NRSI here -->

&nbsp;<!----------------------------------------------------------------------->

@@

In terms of the central research of this thesis, RCTs of statin therapy can be used to provide indirect evidence for the effect of reducing blood LDL-c levels on dementia risk. However, RCTs may be infeasible if the outcome of interest is one with a long prodomal period, such as dementia (see Section \@ref(underlying-pathologies)), as they would require extremely long and costly follow-up.[@ritchie2015] It is no surprise then that the two previous trials providing evidence on the effect of statins on dementia risk, identified by a recent Cochrane review,[@mcguinness2016] are in fact trials of statins for the prevention of coronary related outcomes.

While being widely cited, these studies have major limitations that reduce their utility as a source of evidence on the effect of statin treatment in assessing the impact of lipid-lowering treatment on dementia risk. Firstly, the criteria used to determine a dementia outcome are unclear. One of the trials, the Prospective Study of Pravastatin in the Elderly (PROSPER) trial,[@trompet2010] reported not on dementia outcomes but on the change in cognitive scores over a mean of 3.2 years. As highlighted in Section \@ref(diagnostic-criteria), a "change in score" alone is insufficient to diagnose a dementia outcome. The second trial, the Medical Research Council/British Health Foundation Protection Study,[@heartprotectionstudycollaborativegroup2002] found no effect of simvastatin on dementia (`r estimate(1, 0.61, 1.65)`), but did not report how the outcome was assessed/recorded within the trial.

Additionally, the two trials did not make any effort to assign an underlying pathology to each case, instead reporting an all-cause dementia outcome. As discussed in Section \@ref(underlying-pathologies), the different underlying pathologies of dementia have different mechanisms of action, and so it is not gauranteed that the effect of statins would be consistent across them.

Both trials were also limited by the relatively short follow-up period examined. This limited follow-up was expected, as the primary aim of both trials was to assess the effect of statins on short-term coronary-related outcomes.[@heartprotectionstudycollaborativegroup2002; @trompet2010] The PROSPER trial had a mean follow-up of 3.2 years, while the MRC/BHF Protection Study estimated risk at 5 years of follow-up. Given the long lag time between non-symptomatic onset of dementia and clinical presentation, it is likely that these durations are insufficient to fully capture the onset of dementia. Finally, as they included only patients at high vascular risk, their generalisability to other settings is limited.[@mcguinness2016]

&nbsp;<!----------------------------------------------------------------------->

@@

Newer methodological approaches, such as Mendelian randomisation (MR),[@daveysmith2014] have also been used to examine the effect of varying lipid levels on dementia risk. Mendelian randomisation attempts to combat the risk of reverse causation and residual confounding inherent to observational studies by using natural random varation in participants genomes.[@greenland2000] In brief, MR uses genetic variants that are both strongly associated with the exposure of interest and are independent from potential confounders to strengthen causal inference.[@daveysmith2014] The analytic method relies on several assumptions about the instrumental variable (IV),[@davies2018] namely that:

@@

Recent MR studies indicated that genetically determined low levels of LDL-c may cause a reduction in AD risk.[@larsson2017; @ostergaard2015] However, the effect was attenuated in sensitivity analysis that exclude the region surrounding the ApoE gene, the strongest known risk factor for Alzheimer's disease.[@kim2009] Inclusion of _ApoE4_ variants invalidates the exclusion restriction criteria (Assumption 3, above), as the risk reduction observed may be driven by variants in this region via a pathway independent of lipid levels. This was supported by further MR studies where _ApoE4_ variants were intentionally excluded.[@benn2017] Despite the increasing number of MR studies examining this topic, no systematic review of this study design as a source of evidence has been performed.

@@

Evidence synthesis is the process of finding and integrating information from several sources to examine a research question.[@donnelly2018] A common type of evidence synthesis is a systematic review, either with or without a meta-analysis.[@chandler2019chapter]

@@

* Inclusion of preprints
* Individual patient data meta-analysis
* Triangulation across evidence sources

@@

The importance of including grey literature, defined as literature not published in peer-reviewed journals,[@paez2017] in systematic reviews is widely acknowledged. Meta-research studies have demonstrated that systematic reviews excluding grey literature sources overestimate the effect of interventions.[@conn2003; @mcauley2000; @hopewell2007] Common, well-accepted forms of grey literature include conference abstracts, government reports and theses.[@paez2017;@lefebvre2019searching]

However, the role of preprints in evidence syntheses is less well defined. Defined by the Committee on Publication Ethics (COPE) as 'scholarly manuscript[s] posted by the author(s) in an openly accessible platform, usually before or in parallel with the peer review process'[@committeeonpublicationethicscope2018], preprints serve several purposes. They are used to establish primacy when submitting to a journal where the peer-review process may take several months.[@vale2016] In addition, the allow for the rapid disseminate research findings, as occurred during the COVID-19 pandemic.[@fraser2020preprinting] Finally, they are a source of information on manuscripts that may not have been accepted elsewhere, helping to combat publication bias or the "file-drawer" effect.[@rosenthal1979]

@@

In this thesis, preprints are considered an important source of evidence, in contrast to previous reviews on this topic. However, as with grey literature,[@mahood2014] there are several logistical problems with carrying out systematic searches in preprint repositories. As such, to enable the inclusion of preprints in the systematic review described in Chapters \@ref(sys-rev-methods-heading), a new tool addressing these issues is presented in Chapter \@ref(sys-rev-tools).

&nbsp;<!----------------------------------------------------------------------->

Individual participant data (IPD) meta-analyses are commonly considered to be the gold standard in evidence synthesis methodology.[@riley2010; @stewart1993] IPD methods seek to obtain the raw data from each study identified in a systematic review, rather than basing the meta-analysis on summary results extracted from the literature.[@riley2010]

In the context of this thesis, if lipids are found to have a causal role in development of dementia, evidence-based preventative strategies would be best informed by identifying the types of individuals who are most likely to receive benefit from treatment with lipid-modifying agents.[@arain2009; @clayton2018; @mccartney2016] However, if primary studies do not present results stratified by covariates of interest, meta-analyses of summary-level data on this topic often have limited ability to examine research questions related to exposure-covariate interactions.[@riley2010] In terms of this thesis, participant age and sex are considered to be of particular interest.[@arain2009; @letenneur1999]

An IPD meta-analysis of lipid levels on dementia outcomes would overcome this limitation of summary-level data, as access to the raw data allows for an analysis that investigates these interactions.[@riley2020] This approach has the added benefit of allowing a common set of inclusion criteria and a common statistical model to be applied across all datasets, potentially eliminating some important sources of heterogeneity.[@stewart2002]

Despite their advantages, IPD meta-analysis are rarely performed.[@tugwell2010] Factors limiting their uptake include the increased time and effort they require when compared to a summary-level analysis, and the low success rate associated with obtaining the raw data.[@nevitt2017; @ventresca2020] The data underlying primary studies are frequently not publicly available,[@alsheikh-ali2011; @federer2018] and the availability of data "available on request from authors" declines rapidly over time.[@vines2014] Several systematic barriers to open data sharing have been identified[@vanpanhuis2014]. Of particular concern for biomedical IPD analyses are legal issues surrounding the sharing of medical data, motivated by concerns around patient privacy.[@wartenberg2010]

I will attempt to obtain the raw data from relevant primary studies identified by the systematic review in Chapters \@ref(sys-rev-methods-heading)/\@ref(sys-rev-results-heading). Any data obtained will be combined with that available from the DPUK portal as part of an IPD meta-analysis in Chapter \@ref(ipd-analysis-heading). This will enable the assessment of the effect of lipids on dementia stratified by key variables such as sex.

@@

As illustrated in Section \@ref(evidence-association), several diverse epidemiological methods have been used to examine the effect of varying blood lipid levels on dementia risk. However, each method is limited by its own biases. Aetiological triangulation is a developing evidence synthesis method that seeks to exploit these inherent differences in study design, and as a result, in biases.[@lawlor2016] If several sources of evidence are available and point towards identical conclusions about an exposure-outcome relationship, and these sources are at risk of unrelated biases, this strengthens our confidence in the result. The ideal scenario is where predicted sources of bias are likely to be in competing directions, one to strengthen the effect of the exposure and the other to attenuate it.[@lawlor2016] As such, triangulating these results can provide a middle-ground between the competing directions of bias. A triangulation approach can also prove useful in helping to prospectively design new studies that are at risk of different sources of bias to those already observed in the published literature.[@munafo2018] However, existing methods for triangulation are limited to a qualitative discussion of the different identified results with respect to their biases,[@lawlor2016] an approach which faces issues in interpretability when assessing many results.

To address this limitation, this thesis seeks to develop and apply a novel quantitative triangulation approach to answer causal questions on the effect of blood lipids on dementia outcomes. All existing evidence, regardless of study design, is first identified and assessed using domain-based tools by the systematic review presented in Chapters \@ref(sys-rev-methods-heading)/\@ref(sys-rev-results-heading). The existing evidence base is then supplemented by two primary analyses, presented in Chapters \@ref(cprd-analysis-heading) and \@ref(ipd-analysis-heading). Finally all existing and new evidence is incorporated into the quantiatative triangulation framework in Chapter \@ref(tri-heading).

&nbsp;<!----------------------------------------------------------------------->

@@

* To create a tool that allows for the inclusion of health related preprints in evidence syntheses in a systematic and reproducible manner
* To review all available evidence across multiple diverse study designs to assess the effect of lipids and lipid regulating agents on dementia risk
* To examine whether there is evidence for an effect of lipid-regulating agents on dementia and related outcomes in a large scale population-based cohort, the Clinical Practice Research Datalink (CPRD)
* To produce evidence on lipid-covariate interactions as part of an IPD meta-analysis 
* To propose a generalised framework for the quantitative triangulation across diverse evidence sources

&nbsp;<!----------------------------------------------------------------------->

@@

_**McGuinness L. A.**, Higgins J. P. T., Walker, V. M., Davies, N. M., Martin, R. M., Coulthard, E., Davey-Smith, G., Kehoe, P. G., and Ben-Shlomo, Y. (2021) Association of lipid-regulating drugs with dementia and related conditions: an observational study of data from the Clinical Practice Research Datalink. medRxiv  [10.1101/2021.10.21.21265131](https://doi.org/10.1101/2021.10.21.21265131)_
